---
title: "**PORTO SEGURO SAFE DRIVER PREDICTION**"
output: html_document
---

### **MAIN OBJECTIVE**

The objective of this document is to explore the dataset from the kaggle competition opened in 2017 by Porto Seguro in Kaggle (<https://www.kaggle.com/competitions/porto-seguro-safe-driver-prediction>). Also is the objective to explore several methods of prediction for target variable.

### **NECESSARY LIBRARIES**

These are all the libraries that must be installed and loaded before executing this notebook.

```{r}
library(knitr)
```

### **DATA LOAD**

First let's check which one is the current working directory and change it if necessary to the one where the data is. Next, let's read the train data.

```{r}
getwd()
setwd("..") # Moves backwards in the current directory
setwd("C://Users//riosa//Documents//datasets//porto_seguro_safe_driver_prediction")
getwd()
porto_train <- read.csv("train.csv")
```
### **EXPLORATORY DATA ANALYSIS**
```{r}
dim(porto_train)
```
The dataset counts with 595212 rows and 59 columns.\
Let's get the names of all the columns.
```{r}
#data.frame(index = 1:dim(porto_train)[2], colnames(porto_train))
```
```{r}
kable(data.frame(index = 1:dim(porto_train)[2], colnames(porto_train)), caption = "All variable names.")
```
Since the dataset has the missing values codified has "-1" and R uses NA for this, 
let's change all the -1 to NA.
For this let's use a new function.
```{r}
naConvert <- function(value) {
  # Gets any value and if is equal to -1 returns NA, any other case returns the
  # same value
  result <- value
  if (!is.na(value) & value == -1) {
    result <- NA
  }
  return(result)
}

# A little testing...
naConvert("hi")
naConvert(5)
naConvert(-1)
naConvert(-1.5)
# Test over a vector
test_sample <- sample(c(1:100, rep(-1, 100)), 25)
test_sample
vapply(X = test_sample, FUN = naConvert, FUN.VALUE = numeric(1))
# Let's take a small sample of the dataset for testing the function
set.seed(8511)
porto_sample <- porto_train[sample(1:595212, 100), sample(1:59, 25)]
# Are observations classified has -1 present in the sample obtained?
for (i in 1:25) {
  # IF -1's are found, a positive number bigger than zero is returned
  cat("Variable ", names(porto_sample)[i], ":", " ", sum(porto_sample[, i] == -1), "\n", sep = "")
}
# Now we know that there are actually observations classified as -1 in the sample
# let's check again the function
# First with the variable ps_car_05_cat
summary(vapply(FUN = naConvert, porto_sample$ps_car_05_cat, FUN.VALUE = numeric(1)))
# Next with the variable Variable ps_reg_03
summary(vapply(FUN = naConvert, porto_sample$ps_reg_03, FUN.VALUE = numeric(1)))
# Variable Variable ps_car_14
summary(vapply(FUN = naConvert, porto_sample$ps_car_14, FUN.VALUE = numeric(1)))
# Variable Variable ps_car_01_cat
summary(vapply(FUN = naConvert, porto_sample$ps_car_01_cat, FUN.VALUE = numeric(1)))
# And last variable ps_car_03_cat
summary(vapply(FUN = naConvert, porto_sample$ps_car_03_cat, FUN.VALUE = numeric(1)))

# Now let's check which observations have -1's for the whole dataset
# Also, let's store the names of the variables with -1's and the quantity of them
# for a future checking
missingCount <- numeric() # Stores the number of missings found (-1's)
varpos <- numeric() # Stores the position of that variable in colnames(dataset)
for (i in colnames(porto_train)) {
  missingSum <- sum(porto_train[, i] == -1)
  cat("Variable ", i, ":", " ", missingSum, "\n", sep = "")
  if (missingSum != 0) {
    missingCount <- c(missingCount, missingSum)
    varpos <- c(varpos, which(i == colnames(porto_train)))
  }
}
# After applying the function naConvert we will check the resulting number of NA's
# with this results

# Process the change
for (i in colnames(porto_train)) {
  porto_train[, i] <- vapply(X = porto_train[, i], FUN = naConvert, FUN.VALUE = 1)
}
# Checking after the conversion
for (i in colnames(porto_train)) {
  cat("Variable ", i, ":", " ", sum(porto_train[, i] == -1, na.rm = T), "\n", sep = "")
}
# A data.frame for checking that the quantity of missing observations is the same
# before and after applying the function
checkFrame <- data.frame(variable = colnames(porto_train)[varpos], missingBefore = missingCount,
                         missingAfter = rep(0, length(varpos)))

for (i in colnames(porto_train)[varpos]) {
  checkFrame[which(checkFrame$variable == i), "missingAfter"] <- summary(porto_train[, i])[7]
}
# Let's add a new row with the totals and a new column with 1 if both columns are equal
# and 0 otherwise
checkFrame[dim(checkFrame)[1] + 1, ] <- c("total", colSums(checkFrame[, c("missingBefore", "missingAfter")]))
checkFrame$equal <- as.numeric(checkFrame$missingBefore == checkFrame$missingAfter)
checkFrame
# A dataframe for checking that the rest of variables remain at zero missings
# after using the function naConvert
checkFrame2 <- data.frame(variable = colnames(porto_train)[-varpos])
checkComplete <- numeric() # Variable to store the results of checking that no new
# NA's were introduced
for (i in colnames(porto_train)[-varpos]) {
  checkComplete <- c(checkComplete, sum(is.na(porto_train[, i])))
}
checkFrame2$missingAfter <- checkComplete
checkFrame2
```
